---
title: "R-INLA Practical: spatial and spatio-temporal disease mapping"
subtitle: "GEOMED 2019 Conference"
author: "Marta Blangiardo and Michela Cameletti"
date: "8/27/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Introduction

In this practical you will extend what already seen in the tutorial and estimate spatial and spatio-temporal disease mapping models. You will study salmonellosis disease in cattles in $n=199$ swiss regions for the years 1991-2008 ($T=18$). The data are included in the workspace \texttt{data\_salmonellosis} which can be loaded in \texttt{R} via 

```{r cars}
load("data_salmonellosis.RData")
```

Typing 
```{r}
ls()
```

you will see several objects: \texttt{data.salm}, \texttt{data.Y}, \texttt{data.offset}, \texttt{map} (this is a `SpatialPolygonsDataFrame` object and will be used for mapping). 

The main data file is  \texttt{data.salm} and typing
```{r}
str(data.salm)
```

you can see that the dataset is a \texttt{data.frame} with $199\times 18 = 3582$ rows and five columns:

- \texttt{Y}: the number of salmonellosis cases for a combination of area/year
- \texttt{offset}: the number of herds in the region
- \texttt{region\_id}: the area identifier
- \texttt{time\_id}: the year identifier 
- \texttt{BEZIRKSNR}: another region ID which matches the one included in the  \texttt{map} (which you will need for mapping). 

Note that this dataset is in **long** format, i.e. each region is repeated 18 times (as there are 18 years); use `View(data.salm)` if you want to explore the data.

The workspace also contains \texttt{data.Y} and \texttt{data.offset} which are the cases and the offset in a \texttt{data.frame} format: 
```{r}
dim(data.Y)
dim(data.offset)
colnames(data.Y)
colnames(data.offset)

n = nrow(data.Y) #n. of areas
T = ncol(data.Y) - 1 #n. of years
```
Note that here the number of rows is given by the regions and the number of columns is given by the years (18) plus a column for the region \texttt{BEZIRKSNR} identifier.

DISCLAIMER: the code provided is one of the many ways to get the results. For advanced `R` users: feel free to ignore it and run your own code up to the INLA model.

# 2. Data preparation and exploration 

Create a new \texttt{data.frame} which includes the region \texttt{BEZIRKSNR} identifier and the risk (probability) of salmonellosis given by Y/offset:
```{r}
prob.salm = data.frame(BEZIRKSNR=data.Y$BEZIRKSNR, 
                        p=data.Y[,1:T]/data.offset[,1:T]) #exclude the BEZIRKSNR col
```

Then transform the risk variables in categorical variables by using the `cut` functions and the following cutoffs: 0, 0.01, 0.5 and 1. Define a new `data.frame`:
```{r}
prob.salm.cat = data.frame(lapply(prob.salm[,2:(T+1)],
                          function(x) (cut(x, c(0,0.01,0.5,1), include.lowest=TRUE)))
                          )
#Include area identifier in the dataframe 
prob.salm.cat$BEZIRKSNR = prob.salm$BEZIRKSNR 
colnames(prob.salm.cat)
```

Finally, plot the probabilities for three years (1991, 2000 and 2008):
```{r}
#Merge prob.salm and the Switzerland shapefile 
map.swiss = map
map.swiss@data = merge(map.swiss@data,
                        prob.salm.cat, by="BEZIRKSNR")
head(map.swiss@data)

spplot(map.swiss,
       c("p.1991","p.2000","p.2008"),
       col.regions= terrain.colors(3),
       as.table=TRUE, 
       main = "Risk of salmonellosis (Y/offset)")
```


# 3. Purely spatial disease mapping model
Now, focussing on year 1991, specify a BYM model as the one described in slide 40 of the lecture and Section 4 of the tutorial. 

The adjacency matrix in the inla format is already available through the object `switzerland.graph` saved in your directory (note that for this case study it is not possible to create the adjacency matrix from the `map.swiss` object - as shown in the tutorial - because there is a problem in the geographic structure of the polygons). 

```{r}
swiss.adj.path = "switzerland.graph"
```

Then define the `formula` and run INLA (use a logGamma(1,0.1) prior for both the precisions):
```{r}
library(INLA)

formula = Y ~ 1 + f(region_id,
                    model="bym",
                    graph=swiss.adj.path,
                    hyper=list(prec.unstruct=list(prior="loggamma",param=c(1,0.1)), 
                               prec.spatial=list(prior="loggamma",param=c(1,0.1))))

output1 = inla(formula,
                family = "poisson",
                data = data.salm[data.salm$time_id==1,], #select just the data of the first year
                offset = log(offset),
                control.compute=list(dic=TRUE))
summary(output1)
```


Now you want to map the spatial random effects. To do this you first bring back $z_i=u_i$ + $v_i$ to the natural scale (from the logarithmic one), compute the posterior mean of $z_i$ and create the corresponding categorical variable:
```{r}
names(output1$marginals.random)
length(output1$marginals.random$region_id)
#Remember that the first n rows include information on z=u+v

#Go back to the natural scale
exp.z = unlist(lapply(output1$marginals.random$region_id[1:n], 
                 function(x) inla.emarginal(exp,x)))
range(exp.z)
#Create a new dataframe 
data.exp.z = data.frame(BEZIRKSNR=data.Y$BEZIRKSNR, 
                            exp.z=exp.z)
#Add also the categorical variable
data.exp.z$exp.z.cat = cut(data.exp.z$exp.z,
                           c(0.88,0.99,1.01,1.2,7.8),
                           include.lowest=T)
head(data.exp.z)
```

Then merge the new dataframe `data.exp.z.c` with the data from the shapefile and finally plot the categorical variable:
```{r}
map.swiss@data =  merge(map.swiss@data,
                        data.exp.z,
                        by="BEZIRKSNR")
spplot(obj=map.swiss, 
       zcol="exp.z.cat",
       col.regions=terrain.colors(4),
       as.table=TRUE)
```

It is also possible to compute the probability $p(z_i>1\mid \mathbf y)$ (or equivalently $p(u_i+v_i>0\mid \mathbf y)$) using the built-in function `inla.pmarginal`:
```{r}
#Compute the probabilities
threshold = 0
prob.z = unlist(lapply(output1$marginals.random$region_id[1:n],
                        function(x) 1-inla.pmarginal(threshold,x)))
#Create a new dataframe
data.prob.z = data.frame(BEZIRKSNR=data.Y$BEZIRKSNR, 
                         prob.z=prob.z)
head(data.prob.z)
```

Then you merge the new dataframe `data.prob.z.cat1` with the data from the shapefile and finally plot the categorical variable with the probabilities:
```{r}
prob.cutoff = c(0,0.2,0.8,1)
map.swiss@data =  merge(map.swiss@data,
                        data.prob.z,
                        by="BEZIRKSNR")
spplot(obj=map.swiss,
       zcol="prob.z",
       col.regions=c("green","white","red"),
       at=prob.cutoff,
       as.table=TRUE,
       main="BYM model - exp(z)")
```

Note the same purely spatial model could be estimated indipendently for all the available years.


# 4. Spatio-temporal disease mapping
Now specify a spatio-temporal dynamic model for the salmonellosis data. The model you want to use is the following (see page 65 of the lecture slides):
\begin{eqnarray*}
y_{it} &\sim& \text{Poisson}(\phi_{it}E_{it})\\ 
\log(\phi_{it}) &=& \beta_0 + u_i + v_i + \lambda_t + \gamma_t\\ 
v_i &\sim& \text{Normal}(0,\sigma^2_v)\\
\mathbf{u} &\sim& \text{ICAR}(\mathbf{W},\sigma^2_u)\\
\gamma_t &\sim& \text{Normal}(0,\sigma^2_{\gamma})\\
\lambda_t &\sim& \text{RW1}(\sigma^2_\lambda) 
\end{eqnarray*}

The model includes, besides the BYM components, two temporal effects (unstructured (`model=iid`) and structured (`model=rw1`)). A logGamma(1,0.1) prior is specified for the four precisions.

To estimate this model you have to create a new temporal identifier as you need to define two time specific components in the model and INLA does not allow to assign more than one $\texttt{f()}$ to any identifier:   

```{r}
data.salm$time_id2 = data.salm$time_id
formula.ST1 = Y ~  1+ f(region_id,
                        model="bym",
                        graph=swiss.adj.path,
                        hyper=list(prec.unstruct=list(prior="loggamma",param=c(1,0.1)), 
                               prec.spatial=list(prior="loggamma",param=c(1,0.1)))) +
                        f(time_id, model="iid",
                          hyper=list(prec=list(prior="loggamma",param=c(1,0.1)))) +
                        f(time_id2, model="rw1",
                          hyper=list(prec=list(prior="loggamma",param=c(1,0.1))))
```

Then run the model
```{r}
output2 = inla(formula.ST1,
                family="poisson",
                data=data.salm,
                offset=log(offset),
                control.compute=list(dic=TRUE))
summary(output2)
```

It is possible to plot the posterior mean of the two temporal effects:
```{r}
par(mfrow=c(1,2))
plot(1991:2008,
     output2$summary.random$time_id$mean,
     type="l", main="IID_time",ylab="Posterior mean",xlab="Year")
plot(1991:2008,
     output2$summary.random$time_id2$mean, 
     type="l", main="RW1_time",ylab="Posterior mean",xlab="Year")
par(mfrow=c(1,1))
```

It would be also possible to compute the **sum** of the two temporal effect as the total temporal trend $\gamma_t+\lambda_t$. To do this you need to use `inla.make.lincombs` function (see for example Section 7.1.1 of the INLA book) to create a linear combination of two random effects.

As done in the previous Section 2., it is also possible to plot for this model the residual spatial random effects ($z_i=u_i+v_i$).

# 5. Spatio-temporal disease mapping with space-time interaction
You will now run a spatio-temporal model with a type I interaction (see page 66 of the lecture slides).  

\begin{eqnarray*}
y_{it} &\sim& \text{Poisson}(\phi_{it}E_{it})\\ 
\log(\phi_{it}) &=& \beta_0 + u_i + v_i + \lambda_t + \gamma_t + \delta_{it}\\ 
v_i &\sim& \text{Normal}(0,\sigma^2_v)\\
\mathbf{u} &\sim& \text{ICAR}(\mathbf{W},\sigma^2_u)\\
\gamma_t &\sim& \text{Normal}(0,\sigma^2_{\gamma})\\
\lambda_t &\sim& \text{RW1}(\sigma^2_\lambda) \\
\delta_{it} &\sim& N(0,\sigma^2_{\delta})
\end{eqnarray*}

A logGamma(1,0.1) prior is specified for the five precisions. To run this model you need to specify another identifier for the interaction, going from 1 to the total number of rows in `data.salm`:
```{r}
#ID
region_time_id = seq(1,nrow(data.salm))

#Formula
formula.intI = Y ~  1+ f(region_id,
                        model="bym",
                        graph=swiss.adj.path,
                        hyper=list(prec.unstruct=list(prior="loggamma",param=c(1,0.1)), 
                               prec.spatial=list(prior="loggamma",param=c(1,0.1)))) +
                        f(time_id, model="iid",
                          hyper=list(prec=list(prior="loggamma",param=c(1,0.1)))) +
                        f(time_id2, model="rw1",
                          hyper=list(prec=list(prior="loggamma",param=c(1,0.1)))) +
                        f(region_time_id, model="iid",
                          hyper=list(prec=list(prior="loggamma",param=c(1,0.1))))
#Run INLA
output3 = inla(formula.intI,
              family="poisson",
              data=data.salm,
              offset=log(offset),
              control.compute=list(dic=TRUE))
summary(output3)
```

It is possible to map the **posterior mean** of the interaction $\delta_{it}$ where $i=1,\dots,n$ and $t=1,\ldots,T$:
```{r}
#Extract the posterior mean (log scale) and create a new data frame
delta.intI = data.frame(delta=output3$summary.random$region_time_id$mean,
                        year=data.salm$time_id,
                        ID.area=data.salm$region_id)
dim(delta.intI)
head(delta.intI)
#Transform into categories
delta.intI$delta.cat = cut(delta.intI$delta,
                           c(-0.75,-0.05,0.05,3.1),
                           include.lowest=T)

#Reshape the dataframe so that the dimension is 199 (n) X 18 (T)
delta.intI.wide = reshape(delta.intI,direction="wide",
                          timevar = "year",
                          idvar = "ID.area",
                          drop = "delta")
dim(delta.intI.wide)
#Remove the first column with the station ID
delta.intI.wide = delta.intI.wide[,-1]
#Change the column names
colnames(delta.intI.wide) = paste("delta.",seq(1991,2008),sep="")
#Add the BEZIRKSNR identifier
delta.intI.wide$BEZIRKSNR = data.Y$BEZIRKSNR
```

For mapping it is necessary to merge the shapefile datatable with the new dataframe `delta.intI.wide`:
```{r}
map.swiss@data = merge(map.swiss@data,
                       delta.intI.wide,
                       by="BEZIRKSNR")
spplot(obj=map.swiss,
       zcol=c("delta.1991","delta.2000","delta.2008"),
       col.regions= terrain.colors(3), 
       main="Type I interaction",
       as.table=TRUE)
```

The same approach could be adopted to map for each available year the posterior probability that $\delta$ exceeds the threshold 0:
```{r}
#Extract the probability
prob.delta.vec = unlist(lapply(output3$marginals.random$region_time_id, 
                        function(x){1 - inla.pmarginal(threshold, x)}))

#Create a new dataframe
prob.delta.intI = data.frame(prob.delta=prob.delta.vec,
                        year=data.salm$time_id,
                        ID.area=data.salm$region_id)

#Reshape the dataframe so that the dimension is 199 (n) X 18 (T)
prob.delta.intI.wide = reshape(prob.delta.intI,direction="wide",
                          timevar = "year",
                          idvar = "ID.area")

#Remove the first column with the station ID
prob.delta.intI.wide = prob.delta.intI.wide[,-1]
#Change the column names
colnames(prob.delta.intI.wide) = paste("deltaprob.",seq(1991,2008),sep="")
#Add the BEZIRKSNR identifier
prob.delta.intI.wide$BEZIRKSNR = data.Y$BEZIRKSNR

#Merge before mapping
map.swiss@data = merge(map.swiss@data, prob.delta.intI.wide, by="BEZIRKSNR")
spplot(obj=map.swiss,
       zcol=c("deltaprob.1991","deltaprob.2000","deltaprob.2008"),
       col.regions=c("green","white", "red"), 
       at=prob.cutoff,
       main="Type I interaction - probability",
       as.table=TRUE)
```


Finally we compare the dic of the two spatio-temporal models (output2: spatio-temporal model with bym + 2 temporal effects, output3: spatio-temporal model with bym and space-time type I interaction)
```{r}
output2$dic$dic
output3$dic$dic
```
It seems that the model with the interaction is preferable.


